Black hole symphony orchestra: Spectroscopy using gravitational wave signals from the inspiral, merger and ringdown of binary black holes

### 2020-11-11 papers/intro_paper/intro_paper.tex
updated introduction, waveform model and bayesian PE sections in the paper

\newcommand{\h}{\mathpzc{h}}
\newcommand{\Hhat}{\hat{\mathpzc{H}}}
\newcommand{\B}{\mathpzc{B}}
\newcommand{\hlm}{\mathpzc{h}_{\ell m}}
\newcommand{\xilm}{\xi_{\ell m}}
\newcommand{\Ylm}{{Y}^{-2}_{\ell m}}
\newcommand{\Y}{{Y}^{-2}}
\newcommand{\hc}{h_\times}
\newcommand{\hp}{h_+}
\newcommand{\Fc}{F_\times}
\newcommand{\Fp}{F_+}
\newcommand{\Mf}{M_f}
\newcommand{\cA}{\mathpzc{A}}
\newcommand{\lm}{_{\ell m}}
\newcommand{\deff}{d_\mathrm{eff}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bzeta}{\bm{\zeta}}
\newcommand{\bs}[1]{\bm{\vec{s}_{#1}}}
\newcommand{\Mo}{M_{\odot}}
\newcommand{\FFe}{\mathrm{FF}_\mathrm{eff}}
\newcommand{\FF}{\mathrm{FF}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\rhoopt}{\rho_\mathrm{opt}}
\newcommand{\rhosubopt}{\rho_\mathrm{subopt}}
\newcommand{\fqnm}{f}
\newcommand{\sigmaqnm}{\sigma}
\newcommand{\n}{\mathbf{n}}
\newcommand*{\skymapscale}{0.5}
\newcommand*{\paramestscale}{0.455}
\newcommand{\df}[1]{\delta f_{\text{#1}}}
\newcommand{\dt}[1]{\delta \tau_{\text{#1}}}
\newcommand{\fngr}[1]{f_{\text{#1}}}
\newcommand{\tngr}[1]{tau_{\text{#1}}}
%\newcommand{\fgr}[1]{f ^{\text{GR}}_{\text{#1}}}
%\newcommand{\tgr}[1]{\tau ^{\text{GR}_{\text{#1}}}

@@

\title{Black hole spectroscopy using gravitational wave signals from the inspiral, merger and ringdown of binary black holes}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
The no-hair conjecture in general relativity states that the properties of a Kerr black hole are completely described by its mass and spin angular momentum. As a consequence, the complex quasi-normal-mode (QNM) frequencies of a binary black hole ringdown can be uniquely determined by the mass and spin of the remnant object. Conversely, measurement of the QNM frequencies could be a independent test of the no-hair conjecture. This paper outlines a test of the no-hair conjecture by measuring the complex QNM frequencies of a binary black hole ringdown within the framework of, for the first time, a spinning inspiral-merger-ringdown waveform model, thereby using the entire signal power and removing dependency on the predicted or estimated start time of the proposed ringdown. We further demonstrate the robustness of the test against modified gravitational wave signals with a ringdown different from what general relativity predicts for Kerr black holes. The method was used to analyse the properties of the merger remnants for the events observed by LIGO-Virgo in the first half of their third observing run (O3a) in the latest LIGO-Virgo publication. In this paper, for the first time, we analyse the gravitational wave events from the first and second LIGO-Virgo observing runs and provide joint constraints with published results from O3a. The joint measurements of the fractional deviations in the frequency, $\df{220} = XX \pm$ and $\dt{220} = YY \pm$ are the strongest constraints yet using this method. Finally, we also present a investigation into possible systematic effects due to an incomplete understanding of the interferometric noise around the gravitational wave event on the results of this test.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo Collaborations~\citep{Virgo} recently announced their catalogue of gravitational wave (GW) observations~\citep{GWTC-2} from the first half of their third observing run (O3a)~\citep{O3reference}. Combined with the confirmed observations from the first and second observing runs~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford, Washington and Livingston, Louisiana~\citep{aasi2015characterization}, and the Advanced Virgo detector in Cascina, Italy~\citep{acernese2014advanced} have now detected more than $40$ GW events from the merger of compact objects like neutron stars and/or black holes (compact binary coalescences or CBCs). Alongside independent claims of detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020}, this has firmly established the field of GW astronomy, five years on from the first ever direct detection of GWs on Earth, on September 14, 2015~\citep{abbott2016observation}.

@@

The LIGO-Virgo collaborations have also released a companion paper detailing their latest results of tests of GR using events from the catalogue, GWTC-2. The results include tests of GW generation and source dynamics, where bounds are placed on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of GW propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. The paper also checks for consistency between different portions of the signal using estimates of final mass and spin~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR.

@@

The O3a TGR paper also details results on tests of black hole ringdown and nature of the remnant object, which is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as the underlying theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{LSC_2016grtests}, and later extended to include overtones in~\citep{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Most of the tests mentioned above focus on analysing the post-merger or late-time ringdown signal in isolation. Second generation ground-based interferometric detectors like Advanced LIGO and Virgo are most sensitive to stellar-mass black hole binaries that merge near the minima of their sensitivity band ($\sim 100$Hz). As a consequence, the remnant object \textit{rings down} in shot-noise dominated higher frequencies, leaving very little signal-to-noise ratio (SNR) in the post-merger signal. Furthermore, the post-merger signal of a BBH coalescence transitions from a non-linear to a linear regime where the description of the signal as a QNM spectrum becomes appropriate~\citep{}. The ringdown start time isn't a clearly defined quantity and has been explored in detail in~\citep{Bhagwat:2017tkm}. In other works~\citep{Carullo:2018gah,Carullo:2019flw}, it has been left as a free parameter to be estimated directly from the data. Uncertainties in estimates of the ringdown start-time, as well as an overall lack of SNR in the post-merger signal, given typical sensitivities of ground-based detectors result in significant statistical uncertainties in the measurement of the QNM frequencies.

@@

An independent approach to black hole spectroscopy, based on a full-signal analysis, was outlined in~\citep{Brito:2018rfr} (henceforth referred to as \paperone). \abhi{describe underlying GR model; parameterisation; advantages over other methods; highlight spin}

Unlike methods that focus on the post-merger signal, it describes a framework in which a complete inspiral-merger-ringdown (IMR) waveform model is used to measure the complex QNM frequencies. \abhi{It is a spinning multipolar effective-one-body waveform calibrated to numerical relativity simulations (SEOBNRv4HM~\citep{Cotesta:2018fcv}})This allows access to the full SNR of the signal, reducing measurement uncertainties. Moreover, the definition of the ringdown start time is built into the model and does not need to be either left as an additional free parameter or fixed using alternate definitions. While \paperone presented the method in the context of a non-spinning waveform model, we extend the analysis to the case with black holes spins in the current paper. All astrophysical black holes are expected to be spinning, and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the source properties.
\par
The rest of this paper is organized as follows. Section~\ref{sec:model} describes our parametrised IMR waveform model. In section~\ref{sec:method}, we define our testing GR framework in which we use that model to measure the complex QNM frequencies in a Bayesian formalism. Then in section~\ref{sec:results}, we demonstrate our method on real GW events, as well as simulated signals. Finally we provide a summary of our results and discuss future developments in section~\ref{sec:discussion}.

@@

As in \paperone, we use an IMR waveform model developed within the EOB formalism. However, in this work we extend the results of \paperone by using a waveform model for spinning, nonprecessing binary BHs. Namely we use as our baseline model the multipolar waveform developed in Ref.~\citep{Cotesta:2018fcv}. In addition to being calibrated to NR simulations, this model also uses information from BH perturbation theory for the merger and ringdown phases. Therefore this waveform model provides a very accurate and faithful semi-analytic description of the inspiral, merger and ringdown phases. Henceforth we will denote this model by SEOBNR for short.

@@

where $\iota$ denotes the inclination angle computed with respect to the direction perpendicular to the orbital plane, $\varphi$ is the azimuthal direction to the observer, ${}_{-\!2}Y_{\ell m}(\theta,\phi)$ are the $-2$ spin-weighted spherical harmonics. The SEOBNR model we employ includes the $(\ell, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$ modes~\cite{Cotesta:2018fcv}. For each $(\ell, m)$, the inspiral-(plunge-)merger-ringdown SEOBNR waveform is schematically given by

@@

The merger-ringdown SEOBNR modes read~\citep{Bohe:2016gbl,Cotesta:2018fcv}

@@

are fixed by imposing that the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell m}(t)$ are of class $C^1$ at $t = t_{\textrm{match}}^{\ell m}$, when matching the merger-ringdown waveform to the inspiral-plunge SEOBNR waveform $h_{\ell m}^\mathrm{inspiral-plunge}(t)$. This allows to write the coefficients $c_{i,c}^{\ell m}$ as~\cite{Cotesta:2018fcv}:

@@

In the SEOBNR model constructed in Ref.~\cite{Cotesta:2018fcv}, the complex frequencies $\sigma_{\ell m 0}$  are expressed in terms of the final BH mass and spin~\cite{Berti:2005ys,Berti:2009kk}, and the latter are related to the BBH's component masses and spins through NR--fitting-formulas obtained in GR~\cite{Taracchini:2013rva,Hofmann:2016yih}. Here instead, in the spirit of what was done in \paperone, we promote the QNM (complex) frequencies to be free parameters of the model, while keeping the inspiral-plunge modes $h_{\ell m}^\mathrm{inspiral-plunge}(t)$ fixed to their GR values (henceforth, pSEOBNR model). 
We note that when leaving $\sigma_{\ell m}$ to vary freely, the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell m}(t)$ will in general also differ from the GR prediction, since those functions depend on the QNM complex frequencies, as can be seen from the expressions for $c_{i,c}^{\ell m}$ and $d_{1,c}^{\ell m}$.

For the specific applications of this paper, we will normally only allow either $\sigma_{220}$ and/or $\sigma_{330}$ to vary, while keeping all the other modes to coincide with GR. However the model can be easily used to measure any QNM complex frequency available in the waveform.

@@

A gravitational waveform from the (quasi-circular) coalescence of two black holes is completely described in GR by a set \textit{intrinsic} parameters, $\bxi$: the masses, $m_1, m_2$ and spins, $\bs1, \bs2$ of the initial binary and a reference time and phase $t_c, \phi_c$ respectively; and a set of \textit{extrinsic} parameters, $\btheta$: the location of the binary given by the right ascension, $\alpha$, declination,$\delta$ and the luminosity distance, $d_L$, and its orientation described through the inclination of the binary $\iota$ and its polarisation $\psi$. Also allowing for additional degrees of freedom $\bzeta$ like the complex QNM frequencies described in section \ref{sec:model}, one can then use a Bayesian formalism to obtain a posterior probability density function on the combined parameter set $\blambda = \{\bxi, \btheta, \bzeta\}$:

\begin{equation}
p(\blambda | d, H, I) \propto P(\blambda | H,I) \mathcal{L}(d | \blambda, H, I)
\end{equation}

where $P(\blambda | H,I)$ is the prior probability density function and $\mathcal{L}(d | \blambda, H, I)$ is the likelihood of obtaining the data for a given choice of the parameter set $\blambda$. We assume the hypothesis $H$ to be our waveform model described in section \ref{sec:model} and $I$ to be any other information assumed during the analysis.

Given the full-dimensional posterior probability density function $p(\blambda | d, H, I)$, we can marginalise over the \emph{nuisance} parameters, to obtain the marginalised posterior probability density function on our QNM parameters $\bzeta$, as:

\begin{equation}
p(\bzeta | d, H, I) = \int p(\blambda | d, H, I) d\bxi d\btheta
\end{equation}

An example of a posterior distribution on $\bzeta$Throughout our analysis, we assume the noise in the detector to be well-approximated by the Gaussian random process with a fixed power spectral density. This allows us to define a Gaussian likelihood function $\mathcal{L}(d | \blambda, H, I)$, which we stochastically sample over using either a Markhov-Chain-Monte-Carlo based algorithm implemented in the LIGO Algorithms Library.

>>>>

\newcommand{\h}{\mathpzc{h}}
\newcommand{\Hhat}{\hat{\mathpzc{H}}}
\newcommand{\B}{\mathpzc{B}}
\newcommand{\hlm}{\mathpzc{h}_{\ell m}}
\newcommand{\xilm}{\xi_{\ell m}}
\newcommand{\Ylm}{{Y}^{-2}_{\ell m}}
\newcommand{\Y}{{Y}^{-2}}
\newcommand{\hc}{h_\times}
\newcommand{\hp}{h_+}
\newcommand{\Fc}{F_\times}
\newcommand{\Fp}{F_+}
\newcommand{\Mf}{M_f}
\newcommand{\cA}{\mathpzc{A}}
\newcommand{\lm}{_{\ell m}}
\newcommand{\deff}{d_\mathrm{eff}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bxigr}{\bm{\xi}_{\text{GR}}}
\newcommand{\bxingr}{\bm{\xi}_{\text{nGR}}}
\newcommand{\bzeta}{\bm{\zeta}}
\newcommand{\bs}[1]{\bm{\vec{s}_{#1}}}
\newcommand{\Mo}{M_{\odot}}
\newcommand{\FFe}{\mathrm{FF}_\mathrm{eff}}
\newcommand{\FF}{\mathrm{FF}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\rhoopt}{\rho_\mathrm{opt}}
\newcommand{\rhosubopt}{\rho_\mathrm{subopt}}
\newcommand{\fqnm}{f}
\newcommand{\sigmaqnm}{\sigma}
\newcommand{\n}{\mathbf{n}}
\newcommand*{\skymapscale}{0.5}
\newcommand*{\paramestscale}{0.455}
\newcommand{\df}[1]{\delta f_{\text{#1}}}
\newcommand{\dtau}[1]{\delta \tau_{\text{#1}}}
\newcommand{\fngr}[1]{f ^{\text{nGR}}_{\text{#1}}}
\newcommand{\taungr}[1]{\tau ^{\text{nGR}}_{\text{#1}}}
\newcommand{\fgr}[1]{f ^{\text{GR}}_{\text{#1}}}
\newcommand{\taugr}[1]{\tau ^{\text{GR}}_{\text{#1}}}
\newcommand{\pSEOB}{\texttt{pSEOBNR}}
\newcommand{\SEOB}{\texttt{SEOBNR}}

@@

\title{Black hole symphony orchestra: Spectroscopy using gravitational wave signals from the inspiral, merger and ringdown of binary black holes}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
The no-hair conjecture in general relativity states that the properties of a Kerr black hole are completely described by its mass and spin angular momentum. As a consequence, the complex quasi-normal-mode (QNM) frequencies of a binary black hole ringdown can be uniquely determined by the mass and spin of the remnant object. Conversely, measurement of the QNM frequencies could be a independent test of the no-hair conjecture. This paper outlines a test of the no-hair conjecture by measuring the complex QNM frequencies of a binary black hole ringdown within the framework of, for the first time, a spinning inspiral-merger-ringdown waveform model, thereby using the entire signal power and removing dependency on the predicted or estimated start time of the proposed ringdown. We further demonstrate the robustness of the test against modified gravitational wave signals with a ringdown different from what general relativity predicts for Kerr black holes. The method was used to analyse the properties of the merger remnants for the events observed by LIGO-Virgo in the first half of their third observing run (O3a) in the latest LIGO-Virgo publication. In this paper, for the first time, we analyse the gravitational wave events from the first and second LIGO-Virgo observing runs and provide joint constraints with published results from O3a. The joint measurements of the fractional deviations in the frequency, $\df{220} = XX \pm$ and $\dtau{220} = YY \pm$ are the strongest constraints yet using this method. Finally, we also present a investigation into possible systematic effects due to an incomplete understanding of the interferometric noise around the gravitational wave event on the results of this test.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo Collaborations~\citep{Virgo} recently announced their catalogue of gravitational wave (GW) observations~\citep{GWTC-2} from the first half of their third observing run (O3a)~\citep{O3reference}. Combined with the confirmed observations from the first and second observing runs~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford, Washington and Livingston, Louisiana~\citep{aasi2015characterization}, and the Advanced Virgo detector in Cascina, Italy~\citep{acernese2014advanced} have now detected more than $50$ GW events from the merger of compact objects like neutron stars and/or black holes, collectively called compact binary coalescences (CBCs). Alongside independent claims of detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020}, this has firmly established the field of GW astronomy, five years on from the first ever direct detection of GWs on Earth, on September 14, 2015~\citep{abbott2016observation}.

@@

The LIGO-Virgo collaborations have also released a companion paper detailing their latest results of tests of GR using events from the catalogues GWTC-2 CITE and GWTC-1 CITE. The results include tests of GW generation and source dynamics, where bounds are placed on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of GW propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. The paper also checks for consistency between different portions of the signal using estimates of predicted mass and spin of the remnant object~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR.

@@

The LIGO-Virgo O3a testing GR paper also details results on tests of black hole ringdown and nature of the remnant object, which is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as the underlying theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{LSC_2016grtests}, and later extended to include overtones in~\citep{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Most of the tests mentioned above focus on analysing the post-merger or late-time ringdown signal in isolation. Second generation ground-based interferometric detectors like Advanced LIGO and Virgo are most sensitive to stellar-mass black hole binaries that merge near the minima of their sensitivity band ($\sim 100$Hz). As a consequence, the remnant object \textit{rings down} in shot-noise dominated higher frequencies, leaving very little signal-to-noise ratio (SNR) in the post-merger signal. Furthermore, the post-merger signal of a BBH coalescence transitions from a non-linear to a linear regime where the description of the signal as a QNM spectrum becomes appropriate~\citep{}. The ringdown start time is not a clearly defined quantity and has been explored in detail in~\citep{Bhagwat:2017tkm}. In other works~\citep{Carullo:2018gah,Carullo:2019flw}, it has been left as a free parameter to be estimated directly from the data. Uncertainties in estimates of the ringdown start-time, as well as an overall lack of SNR in the post-merger signal, given typical sensitivities of ground-based detectors result in significant statistical uncertainties in the measurement of the QNM frequencies.

@@

An independent approach to black hole spectroscopy, based on a full-signal analysis, was outlined in~\citep{Brito:2018rfr} (henceforth referred to as \paperone). Unlike methods that focus on the post-merger signal, it describes a framework in which a complete inspiral-merger-ringdown (IMR) waveform model is used to measure the complex QNM frequencies. \abhi{It is a spinning multipolar effective-one-body waveform calibrated to numerical relativity simulations (SEOBNRv4HM~\citep{Cotesta:2018fcv}})This allows access to the full SNR of the signal, reducing measurement uncertainties. Moreover, the definition of the ringdown start time is built into the model and does not need to be either left as an additional free parameter or fixed using alternate definitions. While \paperone presented the method in the context of a non-spinning waveform model, we extend the analysis to the case with black holes spins in the current paper. All astrophysical black holes are expected to be spinning, and ignoring effects of spin have been shown to introduce systematic biases in the measurement of the source properties CITE \cite{paper_showing_systematics_from_ignoring_spin}.
The rest of this paper is organized as follows. Section~\ref{sec:model} describes our parametrised IMR waveform model. In section~\ref{sec:method}, we define our testing GR framework in which we use that model to measure the complex QNM frequencies within a Bayesian formalism. Then in section~\ref{sec:results}, we demonstrate our method on real GW events, as well as simulated signals. Finally we provide a summary of our results and discuss future developments in section~\ref{sec:discussion}.

@@

As in \paperone, we use an IMR waveform model developed within the EOB formalism. However, in this work we extend the results of \paperone by using a waveform model for spinning, nonprecessing binary BHs. Namely we use as our baseline model the multipolar waveform developed in Ref.~\citep{Cotesta:2018fcv}. In addition to being calibrated to NR simulations, this model also uses information from BH perturbation theory for the merger and ringdown phases. Therefore this waveform model provides a very accurate and faithful semi-analytic description of the inspiral, merger and ringdown phases. Henceforth we will denote this model by $\SEOB$ for short.

@@

where $\iota$ denotes the inclination angle computed with respect to the direction perpendicular to the orbital plane, $\varphi$ is the azimuthal direction to the observer, ${}_{-\!2}Y_{\ell m}(\theta,\phi)$ are the $-2$ spin-weighted spherical harmonics. The $\SEOB$ model we employ includes the $(\ell, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$ modes~\cite{Cotesta:2018fcv}. For each $(\ell, m)$, the inspiral-(plunge-)merger-ringdown $\SEOB$ waveform is schematically given by

@@

The merger-ringdown $\SEOB$ modes read~\citep{Bohe:2016gbl,Cotesta:2018fcv}

@@

are fixed by imposing that the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell m}(t)$ are of class $C^1$ at $t = t_{\textrm{match}}^{\ell m}$, when matching the merger-ringdown waveform to the inspiral-plunge $\SEOB$ waveform $h_{\ell m}^\mathrm{inspiral-plunge}(t)$. This allows to write the coefficients $c_{i,c}^{\ell m}$ as~\cite{Cotesta:2018fcv}:

@@

In the $\SEOB$ model constructed in Ref.~\cite{Cotesta:2018fcv}, the complex frequencies $\sigma_{\ell m 0}$  are expressed in terms of the final BH mass and spin~\cite{Berti:2005ys,Berti:2009kk}, and the latter are related to the BBH's component masses and spins through NR--fitting-formulas obtained in GR~\cite{Taracchini:2013rva,Hofmann:2016yih}. Here instead, in the spirit of what was done in \paperone, we promote the QNM (complex) frequencies to be free parameters of the model, while keeping the inspiral-plunge modes $h_{\ell m}^\mathrm{inspiral-plunge}(t)$ fixed to their GR values (henceforth, $\pSEOB$ model). The $\pSEOB$ model uses a parameterised version of the $\SEOB$ model where the frequency and the damping time of the ${\ell m 0}$ mode, i.e, $(f_{\ell m 0}, \tau _{\ell m 0})$ is defined through fractional deviations from the corresponding GR values, i.e., the predictions of the QNM frequencies obtained using the fits~\cite{Taracchini:2013rva,Hofmann:2016yih}:

\begin{eqnarray}
f_{\ell m 0}^{\text{nGR}} &:=& f_{\ell m 0}^{\text{GR}} (1 + \delta f_{\ell m 0}) \\
\tau _{\ell m 0}^{\text{nGR}} &:=& \tau _{\ell m 0}^{\text{GR}} (1 + \delta \tau_{\ell m 0})
\end{eqnarray}

where $(\delta f_{\ell m 0},\delta \tau_{\ell m 0})$ are the fractional deviations in the frequency and damping time, respectively. We note that when leaving $\sigma_{\ell m}$ to vary freely, the functions $\tilde{A}_{\ell m}(t)$ and $\tilde{\phi}_{\ell m}(t)$ will in general also differ from the GR prediction, since those functions depend on the QNM complex frequencies, as can be seen from the expressions for $c_{i,c}^{\ell m}$ and $d_{1,c}^{\ell m}$. \footnote{For the results discussed in this paper, we restrict ourselves to the $(\ell m) = (2,2)$ and $(3,3)$ modes. Most of the high-mass events, for which this test is most appropriate, observed by LIGO-Virgo are consistent with nearly-equal-mass face-on/off binaries for which power in the subdominant modes is not enough to attempt to measure more than two QNM frequencies at once. Hence, all the other $(\ell m)$ modes are fixed at their GR predictions.}

@@

A gravitational wave signal from the (quasi-circular) coalescence of two black holes is completely described in GR by a fixed set of parameters, $\bxigr$. These can be grouped into the \emph{intrinsic} parameters: the masses, $m_1, m_2$ and spins, $\bs1, \bs2$ of the initial binary and a reference time and phase $t_c, \phi_c$ respectively; and the \emph{extrinsic} parameters: the location of the binary given by the right ascension, $\alpha$, declination,$\delta$ and the luminosity distance, $d_L$, and its orientation described through the inclination of the binary $\iota$ and its polarisation $\psi$. The parametrised model described in ~\ref{sec:model}, $\pSEOB$, introduces an additional set of non-GR parameters, $\bxingr$, which essentially consist of the fractional deviations in the frequency and damping time corresponding to each $(\ell,m)$ QNM present in the GR waveform model $\SEOB$. One then proceeds to use the Bayes theorem to obtain the \emph{posterior} probability distribution on $\blambda = \{\bxigr, \bxingr\}$, given a hypothesis $\mathcal{H}$ and any other information used in the analysis $I$:

\begin{equation}
P(\blambda | d, \mathcal{H}, I) = \frac{P(\blambda | \mathcal{H}, I) \, \mathcal{L}(d | \blambda, \mathcal{H}, I)}{P(d|\mathcal{H}, I)},
\label{eq:Bayes_theorem}
\end{equation}
where $P(\blambda | \mathcal{H}, I)$ is the \emph{prior} probability distribution, and $\mathcal{L}(d | \blambda, \mathcal{H}, I)$ is called the \emph{likelihood} function. The denominator is a normalization constant $P(d|\mathcal{H}, I) := \int P(\blambda | \mathcal{H}, I) \, \mathcal{L}(d | \blambda, \mathcal{H}, I) \, d\blambda$, called the marginal likelihood, or the \emph{evidence} of the hypothesis $\mathcal{H}$. In this case, our hypothesis $\mathcal{H}$ is that the data contains a GW signal that is described by the $\pSEOB$ waveform model $h(\blambda)$  and stationary Gaussian noise described by a power spectral density $S_n(f)$. The likelihood function can consequently be defined as:

\begin{equation}
\mathcal{L}(d | \blambda, \mathcal{H}, I) \propto \exp\big[-\frac{1}{2} \langle d - h(\blambda) \, | \, d -h(\blambda) \rangle \big],
\label{eq:likelihood}\end{equation}
where $\langle . | . \rangle$ is the following noise-weighted inner product:
\begin{equation}
\langle A | B \rangle := \int_{f_\mathrm{low}} ^{f_\mathrm{high}} df \frac{\tilde{A}^*(f)\tilde{B}(f) + \tilde{A}(f)\tilde{B}^*(f)}{S_n(f)}.
\label{eq:nwip}
\end{equation}
$\tilde{A}(f)$ denotes the Fourier transform of $A(t)$ and a $^*$ denotes complex conjugation. The limits of integration ${f_\mathrm{low}}$ and ${f_\mathrm{high}}$ define the bandwidth of the sensitivity of the gravitational wave detector. We usually assume ${f_\mathrm{high}}$ to be the Nyquist frequency whereas${f_\mathrm{low}}$ is dictated by the response of the gravitational detector to low-frequency seismic noise. Owing to the large dimensionality of the parameter set $\blambda$, the posterior distribution $P(\blambda | d, \mathcal{H}, I)$ in equation~(\ref{eq:Bayes_theorem}) is computed by stochastically sampling the parameter space using techniques such as Markov-Chain Monte-Carlo~\cite{xxx} or Nested Sampling~\cite{xxx}. For this paper, we use the lalinferencemcmc~\cite{xx} and bilby codes that provide an implementation of the parallely tempered MCMC and Nested Sampling algorithms respectively, for computing the posterior distributions.

Given the full-dimensional posterior probability density function $P(\blambda | d, \mathcal{H}, I)$, we can marginalise over the \emph{nuisance} parameters, to obtain the marginalised posterior probability density function over the QNM parameters $\bxingr$:

\begin{equation}
P(\bxingr | d, \mathcal{H}, I)= \int P(\blambda | d, \mathcal{H}, I) d\bxigr
\end{equation}

For most of the results discussed in this paper, we would restrict ourselves to the $(\ell m) = (2,2)$ and/or $(3,3)$ modes. In those cases we would assume $\bxingr = \{\df{220},\dtau{220}\}$ and/or  $ \{\df{330},\dtau{330}\}$, and all the other $(\ell m)$ modes are fixed at their GR predictions, i.e., $\delta f_{\ell m 0}) = \delta \tau_{\ell m 0} = 0$. This is because, for most of the high-mass events we find most appropriate for this test, the LIGO-Virgo observations are consistent with nearly-equal-mass face-on/off black hole binaries for which power in the subdominant modes is not enough to attempt to measure more than two QNM frequencies at once.
